---
title: "Interacting with the IPUMS API"
author: "Institute for Social Research and Data Innovation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interacting with the IPUMS API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, results="hide"}
library(vcr)

vcr_dir <- "fixtures"

have_api_access <- TRUE

if (!nzchar(Sys.getenv("IPUMS_API_KEY"))) {
  if (dir.exists(vcr_dir) && length(dir(vcr_dir)) > 0) {
    # Fake API token to fool ipumsr API functions
    Sys.setenv("IPUMS_API_KEY" = "foobar")
  } else {
    # If there are no mock files nor API token, can't run API tests
    have_api_access <- FALSE
  }
}

vcr_configure(
  filter_sensitive_data = list(
    "<<<IPUMS_API_KEY>>>" = Sys.getenv("IPUMS_API_KEY")
  ),
  write_disk_path = vcr_dir,
  dir = vcr_dir
)

check_cassette_names()

modify_ready_extract_cassette_file <- function(cassette_file_name,
                                               fixture_path = NULL,
                                               n_requests = 1) {
  fixture_path <- fixture_path %||% vcr::vcr_test_path("fixtures")
  
  ready_extract_cassette_file <- file.path(
    fixture_path, cassette_file_name
  )
  
  ready_lines <- readLines(ready_extract_cassette_file)
  request_lines <- which(grepl("^- request:", ready_lines))
  
  start_line <- request_lines[length(request_lines) - n_requests + 1]
  
  writeLines(
    c(
      ready_lines[[1]],
      ready_lines[start_line:length(ready_lines)]
    ),
    con = ready_extract_cassette_file
  )
}
```

## Overview

The IPUMS API supports two asset types, both of which are supported by ipumsr:

-   **IPUMS extract** endpoints can be used to submit extract requests
for processing and download completed extract files.

-   **IPUMS metadata** endpoints can be used to discover and explore
available IPUMS data as well as retrieve codes, names, and other
extract parameters necessary to form extract requests.

Use of the IPUMS API enables the adoption of a programmatic workflow
that can help users to:

-   Precisely reproduce the specifications of previous extract requests
-   Easily revise and resubmit previous extract requests to update
analyses with new data
-   Save extract request definitions that can be shared with others
without violating IPUMS conditions
-   Integrate the extract download process with functions to load data
into R
-   Quickly identify and explore available IPUMS data sources

The basic workflow for interacting with the IPUMS API is as follows:

1.  [Define](#define) the parameters of an extract request
2.  [Submit](#submit) the extract request to the IPUMS API
3.  [Check](#check) extract request status
4.  [Download](#download) a completed extract

Before getting started, we'll load the necessary packages for the
examples in this vignette:

```{r, message=FALSE}
library(ipumsr)
library(dplyr)
library(purrr)
```

### API Availability

IPUMS **extract** support is currently available via API for the
following collections:

-   IPUMS USA
-   IPUMS CPS
-   IPUMS International
-   IPUMS NHGIS

IPUMS **metadata** support is currently available via API for the
following collections:

-   IPUMS NHGIS

API support will continue to be added for more collections in the
future. You can check general API availability for all IPUMS collections
with `ipums_data_collections()`.

```{r}
ipums_data_collections()
```

### Note: New features currently in development

The tools in ipumsr may not necessarily support all the functionality
currently supported by the IPUMS API. See the [API
documentation](https://developer.ipums.org/docs/apiprogram/) for more
information about its latest features. Furthermore, the API tools in
ipumsr are in active development, and some functionality may not yet be
stable.

## Set up your API key

To interact with the IPUMS API, you'll need to register for access with
the IPUMS project you'll be using. If you have not yet registered, you
can find links to register for each of the API-supported IPUMS
collections below:

-   [IPUMS
USA](https://uma.pop.umn.edu/usa/user/new?return_url=https%3A%2F%2Fusa.ipums.org%2Fusa-action%2Fmenu)
-   [IPUMS
CPS](https://uma.pop.umn.edu/cps/user/new?return_url=https%3A%2F%2Fcps.ipums.org%2Fcps-action%2Fmenu)
-   [IPUMS
International](https://uma.pop.umn.edu/ipumsi/user/new?return_url=https%3A%2F%2Fipumsi.ipums.org%2Fipumsi-action%2Fmenu)
-   [IPUMS
NHGIS](https://uma.pop.umn.edu/nhgis/user/new?return_url=https%3A%2F%2Fnhgis.ipums.org%2Fnhgis-action%2Fmenu)

Once you're registered, you'll be able to [create an API
key](https://account.ipums.org/api_keys).

By default, ipumsr API functions assume that your key is stored in the
`IPUMS_API_KEY` environment variable. You can also provide your key
directly to these functions, but storing it in an environment variable
saves you some typing and helps prevent you from inadvertently sharing
your key with others (for instance, on GitHub).

You can save your API key to the `IPUMS_API_KEY` environment variable
with `set_ipums_api_key()`. To save your key for use in future sessions,
set `save = TRUE`. This will add your API key to your `.Renviron` file
in your user home directory.

```{r, eval=FALSE}
# Save key in .Renviron for use across sessions
set_ipums_api_key("paste-your-key-here", save = TRUE)
```

The rest of this vignette assumes you have obtained an API key and
stored it in the `IPUMS_API_KEY` environment variable.

## Define an extract request {#define}

Each IPUMS collection has its own extract definition function that is
used to specify the parameters of a new extract request from scratch.
These functions take the form `define_extract_*()`:

-   `define_extract_usa()`
-   `define_extract_cps()`
-   `define_extract_ipumsi()`
-   `define_extract_nhgis()`

Alternatively, use `define_extract_from_json()` to load an extract
definition from an external JSON file (see 
[*share an extract definition*](#share)).

When you define an extract request, you can specify the data to be
included in the extract and indicate the desired format and layout.

For instance, the following defines a simple extract request for the `AGE`,
`SEX`, `RACE`, `STATEFIP`, and `MARST` variables from the 2018 and 2019 American
Community Survey (ACS):

```{r}
usa_extract_definition <- define_extract_usa(
  description = "USA extract for API vignette",
  samples = c("us2018a", "us2019a"),
  variables = c("AGE", "SEX", "RACE", "STATEFIP", "MARST")
)

usa_extract_definition
```

You can include more detailed specifications for specific variables using
the `var_spec()` helper:

```{r}
define_extract_usa(
  description = "USA extract for API vignette",
  samples = c("us2018a", "us2019a"),
  variables = list(
    var_spec("AGE", attached_characteristics = "mother"),
    var_spec("SEX", case_selections = "1")
  )
)
```

Similarly, use `ds_spec()` and `tst_spec()` helpers to define an
NHGIS extract request. For instance, to get tables P1 and P2
from the 2010 SF1a census file at the state level, use the `datasets` argument:

```{r}
nhgis_extract_definition <- define_extract_nhgis(
  description = "NHGIS extract for API vignette",
  datasets = ds_spec(
    "2010_SF1a",
    data_tables = c("P1", "P2"),
    geog_levels = "state"
  )
)

nhgis_extract_definition
```

Or to obtain a state-level time series table for persons by age:

```{r}
define_extract_nhgis(
  description = "NHGIS extract for API vignette",
  time_series_tables = tst_spec("CW3", geog_levels = "state"),
  shapefiles = "us_state_2010_tl2010"
)
```

See the documentation for a particular `define_extract_*()`
function to learn more about other available options when defining an extract.

### Extract request objects

`define_extract_` functions always produce an object that inherits from
the `ipums_extract` class, which can be handled by other API functions
(see `?ipums_extract`). Furthermore, these objects will have a subclass
for the particular collection with which they are associated.

```{r}
class(usa_extract_definition)

class(nhgis_extract_definition)
```

While `ipums_extract` objects for different collections have slightly
different structures, they all support the same API workflow.

The specifications for a given extract request object can be accessed
using standard list-indexing syntax.

```{r}
names(usa_extract_definition$samples)

nhgis_extract_definition$datasets[["2010_SF1a"]]
```

`ipums_extract` objects also contain information about the extract
request's status. This extract hasn't yet been submitted, and therefore
has no extract request number:

```{r}
usa_extract_definition$status

usa_extract_definition$number
```

An extract request will receive a number once it has been submitted to
the API for processing.

### Metadata

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("ipums-metadata-api")
```

To build a valid extract request, you need to know the names and codes
used for the data sources and variables you want to include.

While the IPUMS metadata API is not yet available for microdata
collections, users can find basic sample ID codes with `get_sample_info()`.
For instance:

```{r}
get_sample_info("ipumsi")
```

For microdata variable names, refer to the relevant collection's online
extract builder:

-   [IPUMS
USA](https://usa.ipums.org/usa-action/variables/group){target="_blank"}
-   [IPUMS
CPS](https://cps.ipums.org/cps-action/variables/group){target="_blank"}
-   [IPUMS
International](https://international.ipums.org/international-action/variables/group){target="_blank"}

IPUMS NHGIS users can find the codes used for datasets, data tables,
time series tables, and more by using `get_metadata_nhgis()`.

NHGIS provides summary metadata for all available data sources of a
given type (e.g. datasets, shapefiles) or detailed metadata for a
specific data source.

To explore summary metadata, use the `type` argument. This produces a
`tibble` of information about each available data source and is best
used to determine the name used by the API to reference a data source of
interest.

```{r}
get_metadata_nhgis(type = "datasets") %>%
  head()
```

To get details on a particular data source, pass its name to the
associated argument of `get_metadata_nhgis()`. Use this to determine the
codes and descriptions of various subfields for a given data source
(e.g. data tables for a given dataset, geographic levels for a dataset
or time series table, etc.).

For instance, we could use this metadata to determine that the tables
included in the extract request above (P1 and P2) correspond to total
population counts and urban/rural population counts, respectively:

```{r}
ds_meta <- get_metadata_nhgis(dataset = "2010_SF1a")

ds_meta$data_tables %>%
  filter(name %in% c("P1", "P2"))
```

```{r, echo=FALSE, results="hide", message=FALSE}
eject_cassette("ipums-metadata-api")
```

## Submit an extract {#submit}

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("submit-and-check-extract")
```

To submit an extract definition for processing, use `submit_extract()`.

If no errors are detected in the extract definition, a submitted extract
request will be returned with its assigned number and status. Storing
the returned object can be useful for checking the extract request's
status later.

```{r}
submitted_usa_extract <- submit_extract(usa_extract_definition)

submitted_nhgis_extract <- submit_extract(nhgis_extract_definition)
```

The extract number will be stored in the returned object:

```{r}
submitted_usa_extract$number

submitted_usa_extract$status
```

## Check the status of an extract {#check}

It may take some time for the IPUMS servers to process your extract
request. To determine if your extract is ready to download, you can
check its latest status with `get_extract_info()`, which will return an
`ipums_extract` object reflecting the requested extract definition with
the most current `status`. The `status` of a submitted extract will be
one of `"queued"`, `"started"`, `"produced"`, `"canceled"`, `"failed"`,
or `"completed"`.

If the extract is complete and ready for download, the `download_links`
field will be populated with any extract files that are available for
download.

```{r}
submitted_usa_extract <- get_extract_info(submitted_usa_extract)

submitted_usa_extract$status

submitted_usa_extract$download_links
```

If you forget to capture the return value of `submit_extract()`, you can
use the `get_last_extract_info()` helper to request the information for
your most recent extract request for a given collection:

```{r, eval=FALSE}
submitted_usa_extract <- get_last_extract_info("usa")
```

You can also request the information for any single extract request by
passing the collection and extract number to `get_extract_info()`. These
can be provided either as a single string of the form
`"collection:number"` or as a length-2 vector: `c(collection, number)`.
Several other API functions also support this syntax. Note that there is
no need to include the padded zeroes that are included in extract file
names (e.g. 00010).

```{r, eval=FALSE}
# These are equivalent:
cps_extract_10 <- get_extract_info("cps:10")
cps_extract_10 <- get_extract_info(c("cps", 10))
```

If you regularly use only a single IPUMS collection, you can save
yourself some typing by setting that collection as your default.
`set_ipums_default_collection()` will save a specified collection to the
value of the `IPUMS_DEFAULT_COLLECTION` environment variable. If you
have a default collection set, API functions will use that collection in
all requests, assuming no other collection is specified.

```{r, eval=FALSE}
set_ipums_default_collection("cps") # Set `save = TRUE` to store across sessions
```

```{r, echo=FALSE, results="hide", message=FALSE}
set_ipums_default_collection("cps")
```

```{r}
# Check the default collection:
Sys.getenv("IPUMS_DEFAULT_COLLECTION")

# Most recent CPS extract:
get_last_extract_info()

# Request info on extract request "cps:10"
get_extract_info(10)

# You can still request other collections as usual:
get_extract_info("usa:10")
```

```{r, echo=FALSE, results="hide", message=FALSE}
eject_cassette("submit-and-check-extract")
```

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("wait-for-extract-usa")

submitted_usa_extract <- wait_for_extract(submitted_usa_extract)

eject_cassette("wait-for-extract-usa")

# Leave an extract request to simulate wait_for_extract() output for USA
modify_ready_extract_cassette_file(
  "wait-for-extract-usa.yml",
  fixture_path = "fixtures"
)
```

## Download an extract {#download}

To be available for download, an extract must have a `"completed"`
status. However, some requests that are `"completed"` may still be
unavailable for download, as extracts expire and are removed from IPUMS
servers after a set period of time (72 hours for microdata collections,
2 weeks for IPUMS NHGIS).

You can check whether an extract is ready for download with
`is_extract_ready()`:

```{r}
is_extract_ready(submitted_usa_extract)
```

This function will also warn you if your extract has expired and needs
to be resubmitted.

If your extract is ready to download, you can use `download_extract()`
to download the extract's data files to your local machine.

This will return the path to the downloaded file(s) required to load the
data into R. For microdata collections, this is the path to the DDI
codebook (.xml) file, which can be used to read the associated data
(contained in a .dat.gz file).

For NHGIS, this will be a path to the .zip archive containing the
requested data files or shapefiles. If both data and shapefiles are
requested, `download_extract()` will return a length-2 vector of paths
to each, with data files first.

```{r, eval=FALSE}
# By default, downloads to your current working directory
ddi_path <- download_extract(submitted_usa_extract)
```

The files produced by download_extract can be passed directly into the
reader functions provided by ipumsr.

```{r, eval=FALSE}
ddi <- read_ipums_ddi(ddi_path)
micro_data <- read_ipums_micro(ddi)
```

If your extract request has not finished processing, you'll receive an
error when attempting to download. To wait until the extract is ready
before attempting to download, use `wait_for_extract()`. If the extract
request is not yet ready, R will poll the API regularly (each interval
will increase by 10 seconds) until processing has completed.

Note that `wait_for_extract()` will tie up your R session until the
extract is ready to download. This may be undesirable behavior for large
extracts that take a long time to produce or when the IPUMS servers are
busy. You can set a maximum wait time and adjust other API polling
parameters in `wait_for_extract()`. See `?wait_for_extract` for details.

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("download-nhgis")

download_dir <- file.path(tempdir(), "ipums-api-vignette-downloads")
dir.create(download_dir)

download_extract(
  wait_for_extract(submitted_nhgis_extract),
  download_dir = download_dir,
  overwrite = TRUE
)

nhgis_path <- list.files(
  "fixtures",
  pattern = "nhgis[0-9]{4}_csv.zip",
  full.names = TRUE
)

eject_cassette("download-nhgis")

# Remove extra `get_extract_info()` requests from fixture for faster processing
# when re-knitting. Must keep 2 requests so we correctly mock the use of both
# wait_for_extract() and download_extract() in this cassette.
modify_ready_extract_cassette_file(
  "download-nhgis.yml",
  fixture_path = "fixtures",
  n_requests = 2
)
```

```{r, eval=FALSE}
downloadable_extract <- wait_for_extract(submitted_nhgis_extract)
nhgis_path <- download_extract(downloadable_extract)
```

```{r}
nhgis_data <- read_nhgis(nhgis_path)

head(nhgis_data)
```

See the collection-specific vignettes for more information about loading
IPUMS data into R.

## Share an extract definition {#share}

One exciting feature enabled by the IPUMS API is the ability to share a
standardized extract definition with other IPUMS users so that they can
create an identical extract request themselves. The terms of use for
most IPUMS collections prohibit the redistribution of IPUMS data, but
don't prohibit sharing data extract definitions.

ipumsr facilitates this type of sharing with `save_extract_as_json()`
and `define_extract_from_json()`, which read and write `ipums_extract`
objects to and from a standardized JSON-formatted file.

```{r, eval=FALSE}
cps_extract_10 <- get_extract_info("cps:10")
save_extract_as_json(cps_extract_10, file = "cps_extract_10.json")
```

At this point, you can send `cps_extract_10.json` to another user to
allow them to create a duplicate `ipums_extract` object, which they can
load and submit to the API themselves.

```{r, eval=FALSE}
clone_of_cps_extract_10 <- define_extract_from_json("cps_extract_10.json")
submitted_cps_extract <- submit_extract(clone_of_cps_extract_10)
```

(Note that the code in the previous chunk assumes that the file is saved
in the current working directory. If it's saved somewhere else, replace
`"cps_extract_10.json"` with the full path to the file.)

## Revise a previous extract

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("revise-extract")
```

It's not uncommon to find that you no longer need certain data or would
like to update a familiar extract definition with different or more
current data. ipumsr includes functions for revising a previous extract
definition, facilitating a "revise and resubmit" workflow.

Let's say you decided you wanted to include the 2020 ACS sample as well
as the `RELATE` variable to the data from a previous IPUMS USA extract.
You can retrieve that extract definition as described above and add new
values with `add_to_extract()`. In general, any parameters that can be
passed to a collection's `define_extract_` function can also be provided
to `add_to_extract()`.

```{r}
old_extract <- get_extract_info("usa:33")

names(old_extract$samples)

names(old_extract$variables)
```

```{r}
new_extract <- add_to_extract(
  old_extract,
  samples = "us2020a",
  variables = "RELATE",
  description = "New extract with 2020 data"
)

names(new_extract$samples)

names(new_extract$variables)
```

The new extract contains the contents of the original extract definition
for USA 33 along with the new sample, variable, and format selections.
For fields that take a single value, like `description`, the new value
will automatically replace the existing one:

```{r}
new_extract$description
```

Note that because the returned `ipums_extract` object represents a new
extract definition, it will have been reset to an unsubmitted status and
its extract number and download links will have been removed.

To submit this updated extract definition, just use `submit_extract()`
again:

```{r, eval=FALSE}
new_extract_submission <- submit_extract(new_extract)
```

On the other hand, if there are values that you no longer want to
include in your extract request, use `remove_from_extract()`:

```{r}
new_extract <- remove_from_extract(new_extract, samples = "us2020a")

names(new_extract$samples)
```

Removing a variable, dataset, or time series table will remove
all its associated specifications:

```{r}
nhgis_extract_definition <- define_extract_nhgis(
  description = "Extract request with multiple TSTs",
  time_series_tables = list(
    tst_spec("A00", "county"),
    tst_spec("AV0", "state")
  )
)

nhgis_extract_definition

remove_from_extract(nhgis_extract_definition, time_series_tables = "A00")
```

```{r, echo=FALSE, results="hide", message=FALSE}
eject_cassette("revise-extract")
```

## Get info on past extracts {#recent}

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("recent-extracts")
```

If you know you made a specific extract definition in the past, but you
can't remember the exact number, you can use
`get_extract_history()` to peruse your recent extract requests for
a particular collection.

By default, this returns your 10 most recent extract requests as a list
of `ipums_extract` objects.

```{r, warning=FALSE}
usa_extracts <- get_extract_history("usa")
```

You can adjust how many requests to retrieve with the `how_many` argument:

```{r}
nhgis_extracts <- get_extract_history("nhgis", how_many = 3)

nhgis_extracts
```

Because this is a list of `ipums_extract` objects, you can operate on
them with the familiar API functions. For instance, to revise your
second-most-recent extract for resubmission:

```{r, eval=FALSE}
new_extract <- add_to_extract(
  nhgis_extracts[[2]],
  shapefiles = "us_nation_2010_tl2010"
)
```

You can also filter your past extracts for those with particular characteristics
using `purrr::keep()`:

```{r}
purrr::keep(usa_extracts, ~ "MARST" %in% names(.x$variables))

purrr::keep(
  nhgis_extracts, 
  ~ "2010_SF1a" %in% names(.x$datasets) && 
    "P2" %in% .x$datasets[["2010_SF1a"]]$data_tables 
)
```

Or use the `purrr::map()` family to browse your past extract definitions:

```{r}
purrr::map_chr(usa_extracts, ~ .x$description)
```

```{r, echo=FALSE, results="hide", message=FALSE}
eject_cassette("recent-extracts")
```

## Putting it all together

Because each of the API functions in the standard workflow returns an
`ipums_extract` object, we can easily pipe them together in one
expression:

```{r, eval=FALSE}
usa_data <- define_extract_usa(
  "USA extract for API vignette",
  c("us2018a", "us2019a"),
  c("AGE", "SEX", "RACE", "STATEFIP")
) %>%
  submit_extract() %>%
  wait_for_extract() %>%
  download_extract() %>%
  read_ipums_micro()
```

Not only does this pipeline allow you to obtain IPUMS data without ever
leaving your R environment, but it also allows you to retain a
reproducible record of your process. This makes it much easier to
document your workflow, collaborate with other researchers, and update
your analysis in the future.
