---
title: "Microdata API Requests"
author: "Institute for Social Research and Data Innovation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microdata API Requests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, results="hide"}
library(vcr)

vcr_dir <- "fixtures"

have_api_access <- TRUE

if (!nzchar(Sys.getenv("IPUMS_API_KEY"))) {
  if (dir.exists(vcr_dir) && length(dir(vcr_dir)) > 0) {
    # Fake API token to fool ipumsr API functions
    Sys.setenv("IPUMS_API_KEY" = "foobar")
  } else {
    # If there are no mock files nor API token, can't run API tests
    have_api_access <- FALSE
  }
}

vcr_configure(
  filter_sensitive_data = list(
    "<<<IPUMS_API_KEY>>>" = Sys.getenv("IPUMS_API_KEY")
  ),
  write_disk_path = vcr_dir,
  dir = vcr_dir
)

check_cassette_names()
```

This vignette details the options available for requesting data from IPUMS
microdata projects via the IPUMS API.

If you haven't yet learned the basics of the IPUMS API workflow, you may want
to start with the [IPUMS API introduction](ipums-api.html). The code below
assumes you have registered and set up your API key as described there.

## Supported microdata collections

IPUMS provides several data collections that are classified as microdata. 
Currently, the following microdata collections are supported by the IPUMS 
API (shown with the codes used to refer to them in ipumsr):

* IPUMS USA (`"usa"`)
* IPUMS CPS (`"cps"`)
* IPUMS International (`"ipumsi"`)

API support will continue to be added for more collections in the future. 
See the [API documentation](https://developer.ipums.org/docs/apiprogram/)
for more information on upcoming additions to the API.

In addition to microdata projects, the IPUMS API also supports IPUMS NHGIS
data. For details about obtaining IPUMS NHGIS data using ipumsr, see
the [NHGIS-specific vignette](ipums-api-nhgis.html).

Before getting started, we'll load ipumsr and 
[dplyr](https://dplyr.tidyverse.org/), which will be helpful for this demo:

```{r, message=FALSE}
library(ipumsr)
library(dplyr)
```

## Basic IPUMS microdata concepts

Every microdata extract definition must contain a set of requested _samples_ and
_variables_.

In an IPUMS microdata collection, a _sample_ refers to a distinct combination 
of records and variables. A record is a set of values that describe 
the characteristics of a single unit of measurement (e.g. a single
person or a single household), and _variables_ define the characteristics 
that were measured.

A single sample can contain multiple record types (e.g. person records, 
household records, or activity records, and more), each of which correspond to 
different units of measurement.

Note that our usage of 
the term "sample" does not correspond perfectly to the statistical sense of a 
subset of individuals from a population. Many IPUMS samples are samples in 
the statistical sense, but some are "full-count" samples, meaning they contain 
all individuals in a population.

## IPUMS microdata metadata (forthcoming)

```{r, echo=FALSE, results="hide", message=FALSE}
insert_cassette("micro-sample-ids")
```

Of course, to request samples and variables, we have to know 
the codes that the API uses to refer to them. For samples, the IPUMS API uses 
special codes that don't appear in the web-based extract builder. For variables,
the API uses the same variable names that appear on the web.

While the IPUMS API does not yet provide a comprehensive set of metadata 
endpoints for IPUMS microdata collections, users can use the `get_sample_info()`
function to identify the codes used to refer to specific samples when 
communicating with the API.

```{r}
cps_samps <- get_sample_info("cps")

head(cps_samps)
```

The values listed in the `name` column correspond to the code that you would 
use to request that sample when creating an extract definition to be submitted 
to the IPUMS API.

We can use basic functions from dplyr to 
filter the metadata to samples of
interest. For instance, to find all IPUMS International samples for Mexico, we
could do the following:

```{r}
ipumsi_samps <- get_sample_info("ipumsi")

ipumsi_samps %>% 
  filter(grepl("Mexico", description))
```

IPUMS intends to add support for accessing variable metadata via API 
in the future. Until then, use the web-based extract builder for a given 
collection to find variable names and availability by sample. 
See the 
[IPUMS API documentation](https://developer.ipums.org/docs/v2/workflows/explore_metadata/microdata-hp/) 
for links to the extract builder for each microdata collection with API support.

Alternatively, if you have made an extract previously through the web 
interface, you can use `get_extract_info()` to identify the variable names it 
includes. See the [IPUMS API introduction](ipums-api.html) for more 
details.

```{r, echo=FALSE, results="hide", message=FALSE}
eject_cassette("micro-sample-ids")
```

## Defining an IPUMS microdata extract request

Each IPUMS collection has its own extract definition function that is
used to specify the parameters of a new extract request from scratch.
These functions take the form `define_extract_*()`. For microdata collections,
we have:

-   IPUMS USA: `define_extract_usa()`
-   IPUMS CPS: `define_extract_cps()`
-   IPUMS International: `define_extract_ipumsi()`

When you define an extract request, you can specify the data to be
included in the extract and indicate the desired format and layout.

While each microdata collection has its own extract definition function,
each uses the same syntax. The examples in this vignette use multiple
collections, but the syntax they demonstrate can be applied to all of the
supported microdata collections.

A simple extract definition needs only to contain the names of the
samples and variables to include in the request:

```{r}
cps_ext <- define_extract_cps(
  description = "Example CPS extract",
  samples = c("cps2018_03s", "cps2019_03s"),
  variables = c("AGE", "SEX", "RACE", "STATEFIP")
)

cps_ext
```

This produces an `ipums_extract` object containing the extract request
specifications that is ready to be submitted to the IPUMS API.

When you request a variable in your extract definition, the resulting data 
extract will include that variable for all requested samples where it is 
available. If you request a variable that is not available for any requested 
samples, the IPUMS API will throw an informative error when you try to submit 
your request.

Beyond just specifying samples and variables, there are several additional 
options available to refine the data 
requested in a microdata extract request.

## Detailed variable specifications

The IPUMS API supports several detailed specification options that can be 
applied to individual variables in an extract request: case selections,
attached characteristics, and data quality flags.

Before we describe each of these options in depth, we'll introduce the syntax
used to add them to your extract definition.

### Syntax

To add any of these options to a variable, we need to 
introduce the `var_spec()` helper function.

`var_spec()` bundles all the selections for a given variable together into a 
single object (in this case, a `var_spec` object):

```{r}
var <- var_spec("SEX", case_selections = "2")

str(var)
```

To include this specification in our extract, we simply provide it to the
`variables` argument of our extract definition. When multiple variables are
included, pass a `list` of `var_spec` objects:

```{r}
define_extract_cps(
  description = "Case selection example",
  samples = c("cps2018_03s", "cps2019_03s"),
  variables = list(
    var_spec("SEX", case_selections = "2"),
    var_spec("AGE", attached_characteristics = "head")
  )
)
```

In fact, if you investigate our original extract object from above, you'll 
notice that the variables have automatically been converted to 
`var_spec` objects, even though they were provided as character vectors:

```{r}
str(cps_ext$variables)
```

So, a `var_spec` object with no additional specifications will produce
the default data for a given variable. That is, the following are equivalent:

```{r, eval=FALSE}
define_extract_cps(
  description = "Example CPS extract",
  samples = "cps2018_03s",
  variables = "AGE"
)

define_extract_cps(
  description = "Example CPS extract",
  samples = "cps2018_03s",
  variables = var_spec("AGE")
)
```

Because all specified variables are converted to `var_spec` objects, you can 
also pass a list where some elements are `var_spec` objects and some are just 
variable names. This is convenient when you only have detailed specifications 
for a subset of variables:

```{r}
define_extract_cps(
  description = "Case selection example",
  samples = c("cps2018_03s", "cps2019_03s"),
  variables = list(
    var_spec("SEX", case_selections = "2"),
    "AGE"
  )
)
```


(Samples are also converted to their own `samp_spec` objects, but as
there currently aren't any additional specifications available for samples,
there is no reason to use anything other than a character vector in the 
`samples` argument.)

Now that we've covered the basic syntax for including detailed variable
specifications, we can describe the available options in more depth.

### Case selections

Case selections allow us to limit the data to those
records that match a particular value on the specified variable.

For instance, the following specification would indicate that
only records with a value of `"27"` (Minnesota) or 
`"19"` (Iowa) for the variable `"STATEFIP"` should be included:

```{r}
var <- var_spec("STATEFIP", case_selections = c("27", "19"))
```

Some variables have versions with both general and detailed coding schemes.
By default, case selections are interpreted to refer to the general codes:

```{r}
var$case_selection_type
```

For variables with detailed versions, you can also select on the detailed codes. 

For instance, the IPUMS USA variable RACE is available in both general and detailed 
versions. If you wanted to limit your extract to persons identifying as 
"Two major races", you could do so by specifying a case selection of `"8"`.
However, if you wanted to limit your extract to only persons identifying as 
"White and Chinese" or "White and Japanese", you would need to specify 
_detailed_ codes `"811"` and `"812"`.

To include case selections for detailed codes, set 
`case_selection_type = "detailed"`:

```{r}
# General case selection is the default
var_spec("RACE", case_selections = "8")

# For detailed case selection, change the `case_selection_type`
var_spec(
  "RACE", 
  case_selections = c("811", "812"), 
  case_selection_type = "detailed"
)
```


As noted above, IPUMS intends to add support for accessing variable metadata via API 
in the future, such that users will be able to query variable coding schemes 
right from their R sessions.
Until then, use the IPUMS web interface for a given collection to 
find general and detailed variable codes for the purposes of case selection. 
See the 
[IPUMS API documentation](https://developer.ipums.org/docs/v2/workflows/explore_metadata/microdata-hp/) 
for relevant links.

By default, case selection on person-level variables produces a data file that 
includes only those individuals who match the specified 
values for the specified variables. It's also possible to use case selection to 
include matching individuals _and_ all other members of their households, using 
the `case_select_who` parameter.

The `case_select_who` parameter must be the same for all case selections in an 
extract, and thus is set at the extract level rather than the `var_spec` level. 
To include all household members of matching individuals,
set `case_select_who = "households"` in the extract definition:

```{r}
define_extract_usa(
  description = "Household level case selection",
  samples = "us2021a",
  variables = var_spec("RACE", case_selections = "8"),
  case_select_who = "households"
)
```

### Attached characteristics

IPUMS allows users to create variables that reflect the characteristics of other 
household members. To do so, use the `attached_characteristics` argument of
`var_spec()`.

For instance, to attach the spouse's SEX value to a record:

```{r}
var_spec("SEX", attached_characteristics = "spouse")
```

This will add a new variable (in this case, SEX_SP) to the output data that 
will contain the sex of a person's spouse (if no such record exists, the 
value will be 0).

Multiple attached characteristics can be attached for a single variable:

```{r}
var_spec("AGE", attached_characteristics = c("mother", "father"))
```

Acceptable values are `"spouse"`, `"mother"`, `"father"`, and `"head"`.

### Data quality flags

Some variables in the IPUMS have been edited for missing, illegible, and 
inconsistent values. Data quality flags indicate which values are edited
or allocated.

To include data quality flags for an individual variable, use the 
`data_quality_flags` argument to `var_spec()`:

```{r}
var_spec("RACE", data_quality_flags = TRUE)
```

This will produce a new variable (QRACE) containing the data quality flag for 
the given variable.

To add data quality flags for all variables that have them, set
`data_quality_flags = TRUE` in your extract definition directly:

```{r}
usa_ext <- define_extract_usa(
  description = "Data quality flags",
  samples = "us2021a",
  variables = list(
    var_spec("RACE", case_selections = "8"),
    var_spec("AGE")
  ),
  data_quality_flags = TRUE
)
```

Each data quality flag corresponds to one or more variables, and the codes for 
each flag vary based on the sample. See the documentation for the IPUMS
collection of interest for more information about data quality flag codes.

## Data structure and file format

By default, microdata extract definitions will request data in a rectangular
structure and fixed-width file format. 

Rectangular data are data where only person records are included, and any 
household-level variables are converted to person-level variables by copying the 
values from the associated household record onto all household members. 

To instead create a hierarchical extract, which includes separate records for 
households and persons, set `data_structure = "hierarchical"` in your extract 
definition.

See the [IPUMS data reading vignette](ipums-read.html#hierarchical-extracts) 
for more information about loading hierarchical data into R.

To request a file format other than fixed-width, adjust the `data_format` 
argument. Note that
while you can request data in a variety of formats (Stata, SPSS, etc.), 
ipumsr's `read_ipums_micro()` function only supports fixed-width and csv files.

## Next steps

Once you have defined an extract request, you can submit the extract for
processing:

```{r, eval=FALSE}
usa_ext_submitted <- submit_extract(usa_ext)
```

The workflow for submitting and monitoring an extract request and downloading
its files when complete is described in the 
[IPUMS API introduction](ipums-api.html).
